---
title: "Initial Analysis"
author: "Akshi Chaudhary"
date: '`r format(Sys.Date())`'
output: rmarkdown::github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Load the required packages

```{r, message= FALSE, warning= FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
```

2. Downloading data from a url 

```{r, message=FALSE, warning=FALSE}
times <- read.csv(url('https://raw.githubusercontent.com/akshi8/University_rankings/v1.1/data/external/timesData.csv'), strip.white=TRUE)
expenditure <- read.csv(url('https://raw.githubusercontent.com/akshi8/University_rankings/v1.1/data/external/education_expenditure_supplementary_data.csv'),  strip.white=TRUE)
```

3. Data overview

> University ranking data from Times Higher ranking:

```{r}
head(times)
```

> Country-wise education expenditure data across public, private institutes by institute types over the years

```{r}
head(expenditure)
```


4. Data cleaning, changing data formats and treating Null values

* We can see missing values for expenditures for many countries in some years, replacing numeric values with 0 
* Also for the hypothesis testing we have to look at the average expenditure by countries in various education institutions

```{r}
colnames(expenditure)[4] <- "y1995"
colnames(expenditure)[5] <- "y2000"
colnames(expenditure)[6] <- "y2005"
colnames(expenditure)[7] <- "y2009"
colnames(expenditure)[8] <- "y2010"
colnames(expenditure)[9] <- "y2011"
```

Replacing Null values with 0
```{r}
expenditure <- expenditure %>% mutate(y1995 = ifelse(is.na(y1995),0,y1995)
                                  ,y2000 = ifelse(is.na(y2000),0,y2000)
                                  ,y2005 = ifelse(is.na(y2005),0,y2005)
                                  ,y2009 = ifelse(is.na(y2009),0,y2009)
                                  ,y2010 = ifelse(is.na(y2010),0,y2010) 
                                  ,y2011 = ifelse(is.na(y2011),0,y2011))
```

* Similarly we have to convert the university scores in each country to numeric, as total_score is not numeric in raw data
```{r}

times$total_score <- as.numeric(times$total_score)
```

* Some country names are different in both data sources so we'll have to keep uniform country names, for summarized data

```{r}

Name_mapping <- c("Ireland", "Korea, Republic of", "United States" )

times["country"] <- str_replace(times$country,pattern = "Republic of Ireland", Name_mapping[1])
times["country"] <- str_replace(times$country,pattern = "South Korea", Name_mapping[2])
times["country"] <- str_replace(times$country,pattern = "United States of America", Name_mapping[3])

```


5. Summarizing data based on the input for hypothesis testing and visualization

* For this the total expenditures through `1995-2011` have been averaged for each institution type 

```{r}
school_exp <- expenditure %>% filter(direct_expenditure_type !='Total') %>% 
  mutate(avg_exp = (y1995+ y2000 +y2005+y2009+y2010+y2011)/6) %>% group_by(country,direct_expenditure_type,institute_type) %>% 
    summarise(total_exp = round(sum(avg_exp),2)) %>% arrange(desc(total_exp))
head(school_exp)
```

* Taking the score of the best ranking institute of each country using Times ranking data
* We are assuming this to be the proxy for ranking the higher education system for each country

```{r}
country_score <- times %>% filter(total_score != '') %>% 
  group_by(country) %>% 
  summarise(best_score = max(total_score)) %>% 
  select(country,best_score) %>% arrange(desc(best_score))
```

* Arranging best_scores for each country from highest to lowest 

```{r}
country_score$country <- factor(country_score$country , levels = country_score$country [order(country_score$best_score)])

head(country_score)
```

6. Overall expenditure trends across countries in various levels of educations

```{r, message= FALSE, warning= FALSE}
school_exp %>% ggplot(aes(direct_expenditure_type,total_exp )) + geom_boxplot(aes(color = institute_type)) + labs(title = 'Expenditure by countries as a percentage of Total GDP on Education', y = 'Percentage of total GDP') + 
  theme(axis.text=element_text(size=8),axis.title=element_text(size=10,face="bold" )) + theme_bw() + theme(legend.position = "bottom") + scale_color_discrete("Institute Type:")

```

* Let's see the plot of countries with top `higher` education system rankings

```{r}
country_score %>% top_n(20) %>% 
  arrange(desc(best_score)) %>% 
  ggplot(aes(x = best_score, y = country)) + geom_point(color ='blue') + 
  labs(title = 'Higher education scores for countries', x = 'Education Score', y ='Country') +   theme(axis.text=element_text(size=8),axis.title=element_text(size=14,face="bold")) +
  theme_bw()

```


7. Hypothesis Testing

* Do countries who spend more in public education system( as part of their GDP) rank higher in global higher education ranking?

Null Hypothesis $H_0$ : Percentage GDP expenditure on public institute has no association with education score
Alternate Hypothesis $H_A$ : Percentage GDP expenditure on public institute affects the education score of a country

* Combine score data with public expenditure data and filter for public education expenditure 

```{r, warning= FALSE}
df <- left_join(school_exp, country_score, by = "country") %>% filter(direct_expenditure_type == "Public")
```

* Many countries in the expenditure data don't have very high ranking institutes and therefore their best scores are missing, let's impute missing best score with `1` as that is the least score of times ranking 


```{r}
df <- df %>% mutate(best_score = ifelse(is.na(best_score),1,best_score))
df %>% arrange(desc(total_exp))
```



> Lets try the Linear model for hypotheis testing

* Lets visualize the variables first

```{r}
df %>% ggplot(aes(total_exp,best_score)) + geom_point(aes(color = institute_type)) + geom_smooth(method = lm, aes(group = institute_type, color = institute_type), alpha = 0.1) + labs(title = 'Education score based on expenditure on Public insititions', y = 'Higher Education Score', x = 'Percent GDP expenditure on public education') + 
  theme(axis.text=element_text(size=8),axis.title=element_text(size=10,face="bold" )) + theme_bw() + theme(legend.position = "bottom") + scale_color_discrete(" Institute Type:")
```


* For this we will apply linear regression on total_exp and best_score to see the association between education ranking and public education expenditure


```{r}
summary(lm(best_score ~ total_exp + institute_type , data = df))
```


> Observations from the plot and and linear model

* The plot shows the highest association of public higher education expenditure with higher educations ranking scores
* The higher education score of a country is best explained by the it's public expenditure on higher education institutes and that should be the case also
* The linear model however does not provide a very concrete evidence to reject the null hypothesis 
* The p - value of for higher education expenditure versus score is on the margin of significance level testing i.e. 0.056
* This could mean while public expenditure is important for good higher education ranking of a country, it is not the only variable to explain it

8. Next steps would testing other factors affecting higher education scores 

> Research funding 
> Male-female ratio

